Case 4

### 2017-05-19 EU-Circle/eu-circle_case-study-1.Rmd
Progress on EU-circle case-study 1

```{r,warning=FALSE}
## check the relationship between wet-day 95-percentile and wet-day mean
plot(-log(0.05)*zoo(mu[,2]),zoo(q95[,1]),pch=19,xlim=c(0,120),ylim=c(0,120),
     main='Wet-day mean v.s. 95th wet percentile',
     xlab=expression(paste(-ln(1-0.95)*mu,' (mm/day)')), ylab=expression(paste(q[95],' (mm/day)')))
points(-log(0.05)*zoo(mu[,1]),zoo(q95[,2]),pch=19,col=rgb(1,0,0,0.2))
points(-log(0.05)*zoo(mu[,3]),zoo(q95[,3]),pch=19,col=rgb(0,0,1,0.2))
points(-log(0.05)*zoo(mu[,4]),zoo(q95[,4]),pch=19,col=rgb(0,1,0,0.2))
points(-log(0.05)*zoo(mu[,5]),zoo(q95[,5]),pch=19,col=rgb(0,0.7,0.7,0.2))
points(-log(0.05)*zoo(mu[,6]),zoo(q95[,6]),pch=19,col=rgb(0.7,0,0.7,0.2))
points(-log(0.05)*zoo(mu[,7]),zoo(q95[,7]),pch=19,col=rgb(0.5,0.5,0.5,0.2))
grid()
legend(0,120,loc(mu),pch=19,col=c('black',rgb(1,0,0,0.2),rgb(0,0,1,0.2),rgb(0,1,0,0.2),rgb(0,0.7,0.7,0.2),rgb(0.7,0,0.7,0.2),rgb(0.5,0.5,0.5,0.2)),bty='n',cex=0.5)
```

@@

```{r, fig.height=8}
dryseason <- function(x,na.rm=TRUE) if (sum(is.finite(x))>1) mean(subset(spell(x,threshold=1),is=2),na.rm=na.rm) else NA
if (!file.exists('case1-dryseason.rda')) {
  z <- aggregate(subset(X,it='mjjas'),year,FUN='dryseason')
  save(z,file='case1-dryseason.rda')
} else load('case1-dryseason.rda')
eof.dry <- EOF(z)
plot(eof.dry)
```

@@

```{r, fig.height=8, warning=FALSE}
ds.t2m <- DS(eof.dry,eof.t2m)
plot(ds.t2m)
```

@@

```{r, fig.height=8, warning=FALSE}
ds.slp <- DS(eof.dry,eof.slp)
plot(ds.slp)
```

@@

```{r, fig.height=8, warning=FALSE}
cca.slp.t2m <- CCA(eof.slp,eof.t2m)
plot(cca.slp.t2m)
```

@@

```{r, fig.height=8, warning=FALSE}
ds.res <- DS(eof.res,eof.t2m)
plot(ds.res)
```

### Dry conditions and wild fire statistics

Use the EOFs describing the dry conditions as predictors and wild fire statistics as predictand to assess the connection

```{r}
# z <- DS(wfire,eof.dry)
```

@@

```{r, warning=FALSE}
if (!file.exists('dse.mld.slp.rda')) {
  dse.mld.slp <- DSensemble.eof(eof.dry,predictor=SLP,it='mjjas',path="~/data/CMIP5.monthly",pattern = "psl_Amon_ens_")
  save(dse.mld.slp,file = 'dse.mld.slp.rda')
} else load('dse.mld.slp.rda')
plot(dse.mld.slp)
```

Follow up with the residuals and the large-scale temperture

```{r, warning=FALSE, eval=FALSE,echo=FALSE}
if (!file.exists('dse.mld.t2m.rda')) {
  dse.mld.t2m <- DSensemble.eof(eof.res,predictor=T2M,it='mjjas',path="~/data/CMIP5.monthly",pattern = "tas_Amon_ens_")
  save(dse.mld.t2m,file = 'dse.mld.t2m.rda')
} else load('dse.mld.t2m.rda')
plot(dse.mld.t2m)
```

Add the the two components to get an estimate of the total change. The `dse`objects are lists which contain `zoo` time series:

```{r, eval=FALSE,echo=FALSE}
dse.mld <- dse.mld.slp
gcmnm <- names(dse.mld)
gcmnm <- gcmnm[-grep('info',gcmnm)]; gcmnm <- gcmnm[-grep('eof',gcmnm)]
for (nm in gcmnm) {
  coredata(dse.mld[[nm]]) <- coredata(dse.mld.slp[[nm]]) + coredata(dse.mld.t2m[[nm]])
}
```

@@

### Events

Model number of evens $n_f$ and their area $A_f$ seperately, a little like precipitation ($\mu$ and $f_w$). The area has an impact on the risk of affecting the number of assets. Base this on historic data and relate to large-scale conditions (predictors). Model as a partly stochastic process where the likelihood depends on an external variable. Poisson model (`GLM`) of $n_f$ and the nature of the $A_f$ distribution dictates the model strategy for the area.

@@

>>>>

```{r,warning=FALSE}

cols <- c(rgb(1,0,0,0.2),rgb(0,0,1,0.2),rgb(0,1,0,0.2),rgb(0,0.7,0.7,0.2),rgb(0.7,0,0.7,0.2),rgb(0.5,0.5,0.5,0.2))
## check the relationship between wet-day 95-percentile and wet-day mean
plot(-log(0.05)*zoo(mu[,1]),zoo(q95[,1]),pch=19,xlim=c(0,120),ylim=c(0,120),
     main='Wet-day mean v.s. 95th wet percentile',
     xlab=expression(paste(-ln(1-0.95)*mu,' (mm/day)')), ylab=expression(paste(q[95],' (mm/day)')))
for (i in 1:dim(mu[2])) points(-log(0.05)*zoo(mu[,i]),zoo(q95[,i]),pch=19,col=cols[i])
grid()
legend(0,120,loc(mu),pch=19,col=c('black',cols[1:dim(mu)[2]]),bty='n',cex=0.5)
```

@@

```{r, fig.height=8}
dryseason <- function(x,na.rm=TRUE) if (sum(is.finite(x))>1) mean(subset(spell(x,threshold=1),is=2),na.rm=na.rm) else NA
if (!file.exists('case1-dryseason.rda')) {
  z <- aggregate(subset(X,it='mjjas'),year,FUN='dryseason')
  save(z,file='case1-dryseason.rda')
} else load('case1-dryseason.rda')
attr(z,'variable') <- 'duration'
attr(z,'longname') <- 'mean dry spell duration'
eof.dry <- EOF(z)
plot(eof.dry,new=FALSE)
```

@@

```{r, fig.height=8, warning=FALSE}
ds.t2m <- DS(eof.dry,eof.t2m)
plot(ds.t2m,new=FALSE)
```

@@

```{r, fig.height=8, warning=FALSE}
ds.slp <- DS(eof.dry,eof.slp)
plot(ds.slp,new=FALSE)
```

@@

```{r, fig.height=8, warning=FALSE}
cca.slp.t2m <- CCA(eof.slp,eof.t2m)
plot(cca.slp.t2m,new=FALSE)
```

@@

```{r, fig.height=8, warning=FALSE}
ds.res <- DS(eof.res,eof.t2m)
plot(ds.res,new=FALSE)
```

The downscaling exercise applied to the residual suggests that there is little to gain from using both SLP and temperature for downscaling the statistics of duration of dry days. Hence, a projection using SLP from GCMs as predictor may capture most of the future change. The temperature will nevertheless still be important for fires, as hotter temperatures are more favouralbe for fires.

@@

```{r, warning=FALSE}
if (!file.exists('dse.mld.slp.rda')) {
  dse.mld.slp <- DSensemble.eof(eof.dry,predictor=SLP,it='mjjas',ip=1:10,path="~/data/CMIP5.monthly",pattern = "psl_Amon_ens_")
  save(dse.mld.slp,file = 'dse.mld.slp.rda')
} else load('dse.mld.slp.rda')
plot(dse.mld.slp,new=FALSE)
grid()
```

Plot maps of changes in the mean dry spell duration

```{r}
 zmap1 <- map(dse.mld.slp,FUN="mean",FUNX='mean',it=c(2000,2009),plot=FALSE)
 zmap2 <- map(dse.mld.slp,FUN="mean",FUNX='mean',it=c(2090,2099),plot=FALSE)
 zmap <- zoo(coredata(zmap2) - coredata(zmap1),order.by=index(zmap2))
 zmap <- attrcp(zmap1,zmap)
 class(zmap) <- class(zmap1)
 attr(zmap,'variable') <- 'mean dry duration'
 attr(zmap,'unit') <- 'days'
 rm('zmap1','zmap2')
 map(zmap,colbar=list(pal='warm'),main='Change in mean duration of dry spells from 2000-2009',new=FALSE)
```

@@

#### Events

Model number of evens $n_f$ and their area $A_f$ seperately, a little like precipitation ($\mu$ and $f_w$). When it comes to number of events, we can look at the total number or the number of events associated with a threshold area - the number of sevre fires.

The area has an impact on the risk of affecting the number of assets. Base this on historic data and relate to large-scale conditions (predictors). Model as a partly stochastic process where the likelihood depends on an external variable. Poisson model (`GLM`) of $n_f$ and the nature of the $A_f$ distribution dictates the model strategy for the area.

@@

```{r}
## Read wild fire data
colnames <- c('year','number','type','department','code-INSEE','municipality','something','DFCI.code','Alert','origin','area')
fires <- read.table('~/Downloads/liste_incendies_ du_09_05_2017.csv',sep=';',col.names=colnames,header=FALSE)
h <- hist(fires$year,col='grey',breaks=1973:2016)
```

Examine the fire event statistics: is it Poisson distributed?

```{r}
breaks <- seq(0,3000,by=250)
lambda <- mean(h$counts)
hist(h$counts,breaks=breaks,freq = FALSE,ylim=c(0,0.02))
lines(seq(0,3000,by=10),dpois(seq(0,3000,by=10),lambda=lambda),lwd=3,col='red')
```

The annual number is not Poisson distributed. Examine the statistics of the area.

```{r}
summary(fires$area)
plot(fires$area,log='y')
```

Compare the fire statistics with the mean duration of dry intervals from ECAD (eof.dry), the temperature and SLP.

```{r}
## Number of annual fires
naf <- zoo(h$counts,order.by=trunc(h$mids))
naf[naf==0] <- NA
index(naf) <- year(naf)
xy <- merge(naf,zoo(eof.dry),all=FALSE)
cal.naf <- data.frame(y = coredata(xy$naf), x1=coredata(xy$x.1), x2=coredata(xy$x.2),
                      x3= coredata(xy$x.3), x4=coredata(xy$x.4))
naf.model <- lm(y ~ x1 + x2 + x3 + x4, data=cal.naf)
z.naf <- zoo(predict(naf.model),order.by=trunc(h$mids))
plot(naf,z.naf)
```

#### Burned area

```{r}
firearea <- data.frame(i=rep(1,length(fires$year)),area=as.numeric(as.character(fires$area)),year=fires$year)
## Only include fires where more than 1 km^2 has burned down
Af <- aggregate(firearea,by=list(firearea$year,firearea$area > 100000),FUN=sum)

## Number of events with Af > 1km^2 per year
plot(Af$Group.1[Af$Group.2],Af$i[Af$Group.2])

## Number of events with Af > 1km^2 per year
plot(Af$Group.1[Af$Group.2],Af$area[Af$Group.2])

## Relationship between the number of events and total area.
col <- rgb((2015 - Af$Group.1[Af$Group.2])/42,0,1- (2015 - Af$Group.1[Af$Group.2])/42,0.4)
plot(Af$i[Af$Group.2],Af$area[Af$Group.2],col=col,pch=19)
## annual total area of fires
ataf <- as.station(zoo(Af$area[Af$Group.2],order.by=Af$Group.1[Af$Group.2]),loc='Southern France',param='area',unit='m^2',
                  longname='burned area')
plot(ataf,new=FALSE)
```

The number of wild fire events appears to have decreased over time and there burned area is characterised by a weak decline in years with moderate fire activity. There are also some exceptional years with large burned area, but the sample of these extreme cases is too small for a proper trend analysis.  The peak years in fire activity may suggest that the fire hazard is sensitive to certain factors or that increases abruptly when some factors exceed a critical threshold state.

The high sensitivity in the wild fire activity can be compared with the predictors if we take ln(Af) rather than using a linear relationship:

```{r}
index(ataf) <- year(ataf)
index(eof.dry) <- year(eof.dry)
eof.dry <- subset(eof.dry,ip=1:4)
ds.ataf.dry <- DS(log(ataf),eof.dry)
ds.ataf.t2m <- DS(log(ataf),eof.t2m)
ds.ataf.slp <- DS(ataf,eof.slp)
```

Examine the connection between the mean dry spell duration and the area of burned forest

```{r, height=8}
plot(ds.ataf.dry,new=FALSE)
```

The regression results suggest a connection between the mean duration of the dry intervals and ln(Af) as the cross-validation correlation is 0.49.

Also examine the connection between high summer temperature and the burned area:

```{r, height=8}
plot(ds.ataf.t2m,new=FALSE)
```

There is also a connection between ln(Af) and the summer temperature, and the cross-validation correlation is 0.4.

The same analysis repeated for the SLP fails to find a connection for ln(Af), but the linear relationship between SLP and Af captures a dependency

```{r, height=8}
plot(ds.ataf.slp,new=FALSE)
```

The connection between SLP and burned area is weaker than for the the mean dry interval duration and summer temperature if we use ln(Af), with a cross-validation correlation of 0.35. It is particularly the peak years which are associated with high-pressure systems (often associated with dry conditions and heatwaves).

The time series suggest an increasing bias over time, which possibly can be explained by more efficient fire service since 1990. This can be checked to some extent by examining the budget of the fire services to see if there have been more efforts and resources devoted to firefighting over time. It is also possible that the vegetation has changed or that the landscape has been altered (e.g. more fields or more paved area) so that the potential for wild fires has been reduced. Also, once there has been a fire, the are is burned down and it takes time for it to recover so that new fires may occur.

### Projections of future wild fire activity

We can combine the projections of the mean dry interval duration with the downscaling result using the same quantity as predictor. The dry duration statistics was downscaled using the SLP as predictor, so the SLP is in the equation.

```{r}
## Convert the projected data into an EOF for the multi-model ensemble
eof.dry.z <- as.eof(dse.mld.slp)
## Use this EOF as input with the downscaling model calibrated with mean dry interval duration to make projections of ln(fire area)
ds.Af <- exp(predict(ds.ataf.dry,newdata=eof.dry.z))
plot(ds.Af,new=FALSE)
```

#### Simultaneos fires

```{r}
fevents <- data.frame(i=rep(1,length(fires$year)),area=as.numeric(as.character(fires$area)),year=fires$year,date=as.Date(fires$Alert))
## Only include fires where more than 1 km^2 has burned down
se <- aggregate(fevents$i,by=list(fevents$date,fevents$area > 100000),FUN=sum)

## Examine the statistical distribution of simultaneous events
hist(se$x, col='grey', main='Number of simultaneous wild fires')

plot(year(se$Group.1),se$x,pch=19,col=rgb(1,0,0,0.2),main='Number of simultaneous wild fires by year',ylab='Simultaneous events',xlab='')
points(year(se$Group.1),se$x,pch=21,col=rgb(0.5,0.5,0.5))
grid()
```

### 2017-05-19 EU-Circle/eu-circle_case-study-4.Rmd
Progress on EU-circle case-study 1

Get data on storm tracks... IMILAST
Correlation between storm tracks and stations.

## HURDAT2

```{r}
url <- 'http://www.aoml.noaa.gov/hrd/hurdat/hurdat2-1851-2016-apr2017.txt'

```

>>>>

Get data on storm tracks. IMILAST data do not cover the region as the analysis is only for storms north of 20N and south of 20S. Tropical cyclones are reported in the HURDAT2 dataset \url{http://www.aoml.noaa.gov/hrd/hurdat/hurdat2-1851-2016-apr2017.txt}, but only for the North-Atlantic. There are some data in the best-track data base \url{http://www.usno.navy.mil/NOOC/nmfc-ph/RSS/jtwc/best_tracks/}.

## Best Track

```{r}
url <- 'http://www.usno.navy.mil/NOOC/nmfc-ph/RSS/jtwc/best_tracks/'
```

Correlation between storm tracks and stations.
